Modelling plant mortality: integrating intra and interspecific approaches

### 2016-01-29 ms/ms.Rnw
Fixed a few typos

<<preload, echo=FALSE>>=
n_inds <- as.character(round(nrow(BCI_model_dataset_true_unique),-2))
n_spp <- length(unique(BCI_model_dataset_true_unique$sp))
proportion_negative <- round(nrow(subset(BCI_model_dataset_true_unique, obs_dbh_dt<0))/nrow(BCI_model_dataset_true_unique)*100)
n_obs_errors <- nrow(BCI_dbh_error_data$discrep)
obs_error_sigma <- round(sd(BCI_dbh_error_data$discrep),2)
@

@@

Here we examine the performance of three functional forms of $\lambda$ to predict observed patterns of tropical tree mortality. The simplest model (eqn: \ref{base_haz}), assumes a single baseline hazard, which is constant irrespective of a plants growth rate. The second model (eqn: \ref{growth_haz} describes a growth-dependent hazard which assumes plants can only die due to a growth related cause. In this model, we make an explicit assumption that the growth-mortality relationship is captured by a negative exponential curve whereby the probability of death is high at low growth rates and as growth increases the probability of dying decays at an exponential rate until an asymptote at zero is reached. This negative exponential relationship has been used in other studies to model growth-dependent mortality \citep[e.g.][]{Kobe:1995tw} and broadly reflects patterns observed in non-parametric growth-mortality models \citep{Wyckoff:2000un, Kunstler:2005bn, Vieilledent:2010fv}. The third model (eqn: \ref{add_haz}, and potentially the most biologically realistic, is the summation of the two previous models. Specifically, this model allows $\lambda$ to decrease exponentially with increasing growth, but, unlike a standard negative exponential, will asymptote at baseline hazard instead of zero. The advantage of this model is that it effectively accounts for two independent causes of mortality: that due to growth and that due to chance (i.e. stochastic events such as windthrow and fire). The parameters of eqn: \ref{add_haz} are also biologically interpretable: $\alpha$ represents the intercept (i.e. sum of both growth dependent and growth independent mortality at zero growth rate) and biologically defines the instantaneous mortality hazard $\lambda$ at zero growth. $\beta$ represents the exponential decay rate or the sensitivity of hazard rates to change with growth rate; and $\gamma$, is the asymptote and defines the baseline hazard. $\alpha$ and $\beta$ are intrinsically linked to growth rate, and as such are growth-dependent parameters. By contrast, $\gamma$ is a growth-independent parameter that accounts for stochastic events (e.g. windfall, fire) that can kill a plant irrespective of its growth rate.
\begin{align}
\lambda &= \overbrace{\gamma}^{\text{baseline hazard}} \label{base_haz}\\
\lambda &= \overbrace{\alpha \, \exp\left(-\beta \, Growth_i \right)}^{\text{Growth-dependent hazard}} \label{growth_haz}\\
\lambda &= \overbrace{\alpha \, \exp\left(-\beta \, Growth_i \right)}^{\text{Growth-dependent hazard}} + 
\overbrace{\gamma}^{\text{Baseline hazard}} \label{add_haz}
\end{align}

A central aim of this study is to identify what is the most predictive parameterisation of $\lambda$. To do this, we used K-fold cross-validation and four iterative stages of model fitting and comparison. Stage 1 involved comparing predictive performance of the three $\lambda$ models described above (eqn: \ref{base_haz,growth_haz,add_haz}). In this stage each model was fitted with fixed parameters (i.e. parameters were constant across species) and then again by allowing parameters to varying by species. This allowed us to determine: 1) whether predictive performance was substantially improved with increasing functional complexity; and 2) the extent to which species hazard rates differ. In stage 2 we used the most predictive model from stage 1 to examine whether dbh growth or basal area growth was a better predictor of plant mortality. In stage 3 we used the best model from stage 2, but without species random effects, to determine what combination of parameter-wood density effects best predicted tree mortality. This stage also allowed us to determine whether wood density primarily affected growth-dependent or growth-independent (i.e. baseline) mortality. Finally, in stage 4, we fit the best model based on the previous three steps to the entire dataset with species random effects on all parameters. We then examined wood density effect sizes and the proportion of species variability that it explained.

@@

We modelled $p_{i,t_1\rightarrow t_2}$ using a complementary log-log link as:
\begin{equation}
\log(-\log(1 - p_{i,t_1\rightarrow t_2})) = \log((t_2-t_1)\,\lambda_i)
\end{equation}.
This parameterisation allowed for additive hazards such as those in \ref{add_haz} and is equivilant to a competiting risks analysis \citep{Kleinbaum:2012tn}.

As we have multiple hypothesized functional forms of $\lambda$ (e.g. eqn: X ), each containing parameters that may or may not vary by species, we outline the fitting procedure of our most complicated model of which all others are nested within:
\begin{equation}
\lambda_i = \overbrace{\alpha_{s[i]} \, \exp\left(-\beta_{s[i]} \, Growth_i \right)}^{\text{Growth dependent}} + 
\overbrace{\gamma_{s[i]}}^{\text{Growth independent}}.
\end{equation}

In this model, $\alpha_s$, $\beta_s$ and $\gamma_s$ varying by species a function of both wood density (measured at species level) and a species random effect:
\begin{equation}\label{eq:submodels}
\begin{aligned}
\alpha_s &= \alpha_{0,s} \, \rho_s^{\alpha_{1}},\\
\beta_s &= \beta_{0,s} \, \rho_s^{\alpha_{1}}, \\
\gamma_s &= \gamma_{0,s} \, \rho_s^{\alpha_{1}}. 
\end{aligned}
\end{equation}
Here $\alpha_{0,s}$,$\beta_{0,s}$,$\gamma_{0,s}$ represents the species-level residual error not explained by wood density for species $s$ within which individual $i$ belongs. $\alpha_{1}$,$\beta_{1}$,$\gamma_{1}$ represents the effect of standardised wood density $\rho_s$ on $\alpha$,$\beta$ and $\gamma$, respectively. The form of submodels in eqn: \ref{eq:submodels} ensures that parameters remain positive. On a log scale these submodels become additive (e.g. $\log \alpha_s = \log \alpha_{0,s} + \alpha_{1} \, \log \rho_s$. We standardised wood density via the transformation: $\rho_s = \frac{\rho}{0.6}$), an operation which is equivalent to centering a log-transformed $\rho$ at a density of 0.6 g cm${^-3}$. This was done for easier interpretation, allowing $\alpha_{1,t}$ to be interpreted as the hazard for a species with wood density of 0.6 cm${^-3}$ (as opposed to species responses at $\rho$ = 0; an biologically unrealistic state for a woody plant).

@@

We examined predictive accuracy of each model by using K-fold cross-validation. This involved randomly partitioning the data into ten folds of X observations, ensuring that each K-fold had at least one observation per species. We then fitted models to nine folds and predicted to the tenth fold. This was repeated until each fold had been used nine times for model fitting, and once as a heldout dataset for predicting. Cross-validation is a typical approach to model evaluation, and assesses a model’s accuracy when extrapolating to data not used in model calibration, thereby indicating the model’s predictive accuracy (Stone:1997tg).

>>>>

<<preload, echo=FALSE>>=
n_inds <- as.character(round(nrow(BCI_model_dataset_true_unique),-2))
n_spp <- length(unique(BCI_model_dataset_true_unique$sp))
proportion_negative <- round(nrow(subset(BCI_model_dataset_true_unique, obs_dbh_dt<0))/nrow(BCI_model_dataset_true_unique)*100)
n_obs_errors <- nrow(BCI_dbh_error_data$discrep)
obs_error_sigma <- round(sd(BCI_dbh_error_data$discrep),2)
n_heldout <- as.character(nrow(BCI_training_sets[[1]]$heldout))
@

@@

Here we examine the performance of three functional forms of $\lambda$ to predict observed patterns of tropical tree mortality. The simplest model (eqn:\ref{base_haz}), assumes a single baseline hazard, which is constant irrespective of a plants growth rate. The second model (eqn: \ref{growth_haz} describes a growth-dependent hazard which assumes plants can only die due to a growth related cause. In this model, we make an explicit assumption that the growth-mortality relationship is captured by a negative exponential curve whereby the probability of death is high at low growth rates and as growth increases the probability of dying decays at an exponential rate until an asymptote at zero is reached. This negative exponential relationship has been used in other studies to model growth-dependent mortality \citep[e.g.][]{Kobe:1995tw} and broadly reflects patterns observed in non-parametric growth-mortality models \citep{Wyckoff:2000un, Kunstler:2005bn, Vieilledent:2010fv}. The third model (eqn:\ref{add_haz}, and potentially the most biologically realistic, is the summation of the two previous models. Specifically, this model allows $\lambda$ to decrease exponentially with increasing growth, but, unlike a standard negative exponential, will asymptote at baseline hazard instead of zero. The advantage of this model is that it effectively accounts for two independent causes of mortality: that due to growth and that due to chance (i.e. stochastic events such as windthrow and fire). The parameters of eqn:\ref{add_haz} are also biologically interpretable: $\alpha$ represents the intercept (i.e. sum of both growth dependent and growth independent mortality at zero growth rate) and biologically defines the instantaneous mortality hazard $\lambda$ at zero growth. $\beta$ represents the exponential decay rate or the sensitivity of hazard rates to change with growth rate; and $\gamma$, is the asymptote and defines the baseline hazard. $\alpha$ and $\beta$ are intrinsically linked to growth rate, and as such are growth-dependent parameters. By contrast, $\gamma$ is a growth-independent parameter that accounts for stochastic events (e.g. windfall, fire) that can kill a plant irrespective of its growth rate.
\begin{align}
\lambda &= \overbrace{\gamma}^{\text{baseline hazard}} \label{base_haz}\\
\lambda &= \overbrace{\alpha \, \exp\left(-\beta \, Growth_i \right)}^{\text{Growth-dependent hazard}} \label{growth_haz}\\
\lambda &= \overbrace{\alpha \, \exp\left(-\beta \, Growth_i \right)}^{\text{Growth-dependent hazard}} + 
\overbrace{\gamma}^{\text{Baseline hazard}} \label{add_haz}
\end{align}

A central aim of this study is to identify what is the most predictive parameterisation of $\lambda$. To do this, we used K-fold cross-validation and four iterative stages of model fitting and comparison. Stage 1 involved comparing predictive performance of the three $\lambda$ models described in equations \ref{base_haz}, \ref{add_haz}. In this stage each model was fitted with fixed parameters (i.e. parameters were constant across species) and then again by allowing parameters to varying by species. This allowed us to determine: 1) whether predictive performance was substantially improved with increasing functional complexity; and 2) the extent to which species hazard rates differ. In stage 2 we used the most predictive model from stage 1 to examine whether dbh growth or basal area growth was a better predictor of plant mortality. In stage 3 we used the best model from stage 2, but without species random effects, to determine what combination of parameter-wood density effects best predicted tree mortality. This stage also allowed us to determine whether wood density primarily affected growth-dependent or growth-independent (i.e. baseline) mortality. Finally, in stage 4, we fit the best model based on the previous three steps to the entire dataset with species random effects on all parameters. We then examined wood density effect sizes and the proportion of species variability that it explained.

@@

We modelled $p_{i,t_1\rightarrow t_2}$ using a complementary log-log link as:
\begin{equation}
\log(-\log(1 - p_{i,t_1\rightarrow t_2})) = \log((t_2-t_1)\,\lambda_i).
\end{equation}
This parameterisation allowed for additive hazards such as those in eqn:\ref{add_haz} and is equivilant to a competiting risks analysis \citep{Kleinbaum:2012tn}.

As we have multiple hypothesized functional forms of $\lambda$ (e.g. eqn:\ref{base_haz}, \ref{add_haz}), each containing parameters that may or may not vary by species, we outline the fitting procedure of our most complicated model of which all others are nested within:
\begin{equation}
\lambda_i = \overbrace{\alpha_{s[i]} \, \exp\left(-\beta_{s[i]} \, Growth_i \right)}^{\text{Growth dependent}} + 
\overbrace{\gamma_{s[i]}}^{\text{Growth independent}}.
\end{equation}

In this model, $\alpha_s$, $\beta_s$ and $\gamma_s$ varying by species a function of both wood density (measured at species level) and a species random effect:
\begin{equation}\label{eq:submodels}
\begin{aligned}
\alpha_s &= \alpha_{0,s} \, \rho_s^{\alpha_{1}},\\
\beta_s &= \beta_{0,s} \, \rho_s^{\alpha_{1}}, \\
\gamma_s &= \gamma_{0,s} \, \rho_s^{\alpha_{1}}. 
\end{aligned}
\end{equation}
Here $\alpha_{0,s}$,$\beta_{0,s}$,$\gamma_{0,s}$ represents the species-level residual error not explained by wood density for species $s$ within which individual $i$ belongs. $\alpha_{1}$,$\beta_{1}$,$\gamma_{1}$ represents the effect of standardised wood density $\rho_s$ on $\alpha$,$\beta$ and $\gamma$, respectively. The form of submodels in eqn:\ref{eq:submodels} ensures that parameters remain positive. On a log scale these submodels become additive (e.g. $\log \alpha_s = \log \alpha_{0,s} + \alpha_{1} \, \log \rho_s$. We standardised wood density via the transformation: $\rho_s = \frac{\rho}{0.6}$), an operation which is equivalent to centering a log-transformed $\rho$ at a density of 0.6 g cm${^-3}$. This was done for easier interpretation, allowing $\alpha_{1,t}$ to be interpreted as the hazard for a species with wood density of 0.6 cm${^-3}$ (as opposed to species responses at $\rho$ = 0; an biologically unrealistic state for a woody plant).

@@

We examined predictive accuracy of each model by using K-fold cross-validation. This involved randomly partitioning the data into ten folds of \Sexpr{n_heldout} observations, ensuring that each K-fold had at least one observation per species. We then fitted models to nine folds and predicted to the tenth fold. This was repeated until each fold had been used nine times for model fitting, and once as a heldout dataset for predicting. Cross-validation is a typical approach to model evaluation, and assesses a model’s accuracy when extrapolating to data not used in model calibration, thereby indicating the model’s predictive accuracy \citep{Stone:1997tg}.
